---
title: "Day 3 R"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
generation <- read.csv(here::here("data", "generation.csv"), stringsAsFactors = F)
imports <- read.csv(here::here("data", "imports.csv"), stringsAsFactors = F)
```

```{r}
head(generation)
```

```{r}
head(imports)
```

```{r}
str(generation)
```

```{r}
str(imports)
```

```{r}
library(lubridate)
generation$datetime <- as_datetime(generation$datetime)
class(generation$datetime)
```

```{r}
imports$datetime <- as_datetime(imports$datetime)
class(imports$datetime)
```
```{r}
head(generation)
```

melting dataframe
melt(df, id.vars = 'id')
```{r}
library(reshape2)
long_gen <- melt(generation, id.vars='datetime',variable.name = "source", value.name = "usage")
head(long_gen)
```

```{r}
long_gen[order(long_gen$datetime)[1:20], ]
```

joining by datetime
```{r}
merged_energy <- merge(generation, imports, by='datetime')
dim(merged_energy)
head(merged_energy)
```

```{r}
long_merged_energy <- melt(merged_energy, id.vars = 'datetime', variable.name = 'source', value.name = 'usage')
head(long_merged_energy)
```

dplyr
```{r}
library(dplyr)
tmp <- select(merged_energy, biogas, biomass, geothermal, solar)
names(tmp)

```

```{r}
tmp <- select(merged_energy, -biogas, -biomass)
names(tmp)
```

ones_of, contains, starts_with, ends_with, matches
```{r}
tmp <- select(merged_energy, contains('hydro'), starts_with('bio'))
names(tmp)
```

```{r}
tmp <- filter(merged_energy, imports > 7000)
nrow(tmp)
```

```{r}
tmp <- filter(merged_energy, imports>7000, natural_gas < 7000)
nrow(tmp)
```

mutate create new variables
```{r}
tmp <- mutate(long_merged_energy, log_usage = log(usage))
head(tmp)
```

summarize reduce to single value
```{r}
summarize(long_merged_energy, total = sum(usage, na.rm = T))
```

```{r}
summarize(long_merged_energy, mean_cons = mean(usage, na.rm=T))
```

pipe %>%
sends result of one funnction to another function
```{r}
long_merged_energy %>% 
  filter(source == "geothermal") %>%
  select(-datetime) %>%
  mutate(log_usage = log(usage)) %>%
  summarize(mean_log_usage = mean(log(usage), na.rm = T))
```

use period to refer to manipulated dataframe
```{r}
merged_energy %>%
  select(-datetime) %>%
  mutate(total_usage = rowSums(., na.rm = T)) %>%
  summarize(total_usage = sum(total_usage, na.rm = T))
```

```{r}
merged_energy %>%
  select(contains("hydro")) %>%
  mutate(total_hydro = rowSums(., na.rm = T)) %>%
  summarise(mean_hydro = mean(total_hydro, na.rm = T))
```

```{r}
long_merged_energy %>%
  group_by(source) %>%
  summarize(sum_usage = sum(usage, na.rm = T))
```

```{r}
head(long_merged_energy)
```
```{r}
head(merged_energy)
```


```{r}
merged_energy %>%
  select(datetime, contains('bio'), contains('hydro')) %>%
  melt(., id.vars='datetime', variable.name = "source", value.name = "usage") %>%
  group_by(source) %>%
  summarize(mean_usage = mean(usage, na.rm = T))
```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```

```{r}

```




---
title: "dplyr Exercises Day 3"
author: "Richard Paquin Morel, adapted from exercises by Christina Maimone"
date: "`r Sys.Date()`"
output: html_document
params:
    answers: FALSE
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_chunk$set(echo=answers, eval=answers,
                      warning = FALSE, message = FALSE,
                      cache = FALSE, tidy = FALSE)
```

## Load the data

Read both California energy datasets. Make sure you use `lubridate` to recode the `datetime` variable. Use `dplyr` to do this.

```{r}
### Answer

```

```{r}
library(dplyr)
library(lubridate)
generation <- read.csv(here::here("data/generation.csv"), 
                       stringsAsFactors = F)
imports <- read.csv(here::here("data/imports.csv"), 
                    stringsAsFactors = F)

generation <- mutate(generation, datetime = as_datetime(datetime))
imports <- mutate(imports, datetime = as_datetime(datetime))
```


## Merge and reshape the data

Using dplyr and pipes, merge the two datasets and then melt the resulting dataframe to make it tidy.

```{asis}
### Answer
```

```{r}
library(reshape2)
long_ca_energy <- generation %>%
  inner_join(imports, by = "datetime") %>% 
  melt(id.vars = "datetime",
       variable.name = "source",
       value.name = "output")
```


## Creating new variables

Create a series of new variables: 

1. `day`, which is the year-month-day, without the hour. The `lubridate` function `as_date` will do this.
2. `log_output`, which is the natural log of the output.
3. **Challenge**: `per_output`, which is the percent of daily output represented by each observation. You will need to use `group_by` and to create a new variable with the total output for the day. (Make sure to use `ungroup()` after this!)

Try to this all in one pipe!

```{asis}
### Answer
```

```{r}
long_ca_energy <- long_ca_energy %>%
  mutate(day = as_date(datetime),
         log_output = log(output)) %>%
  group_by(day) %>%
  mutate(total_daily_output = sum(output, na.rm = T)) %>% 
  ungroup() %>% 
  mutate(per_output = output/total_daily_output)
```

## Summarizing and analyzing data

1. Which source has the greatest mean output by hour? (Hint: Use the dplyr verb `arrange(desc(variable))` to order the data frame so that the largest value of `variable` is first. Don't use `desc` and it arranges in ascending order.) Which has the least?
2. Which source has the greatest mean output by day? Which has the least?
3. Which sources has the greatest variance in usage over the course of a dataset? Which has the least?

```{asis}
### Answer
```

```{r}
long_ca_energy %>% 
  group_by(source) %>% 
  summarize(mean_hourly = mean(output, na.rm = T)) %>% 
  arrange(desc(mean_hourly))

long_ca_energy %>% 
  group_by(source) %>% 
  summarize(mean_hourly = mean(output, na.rm = T)) %>% 
  arrange(mean_hourly)
```

```{r}
long_ca_energy %>% 
  group_by(day, source) %>% 
  summarize(mean_daily = mean(output, na.rm = T)) %>% 
  arrange(desc(mean_daily))

long_ca_energy %>% 
  group_by(day, source) %>% 
  summarize(mean_daily = mean(output, na.rm = T)) %>% 
  arrange(mean_daily)
```

```{r}
long_ca_energy %>% 
  group_by(source) %>% 
  summarize(sd_output = sd(output, na.rm = T)) %>% 
  arrange(desc(sd_output))

long_ca_energy %>% 
  group_by(source) %>% 
  summarize(sd_output = sd(output, na.rm = T)) %>% 
  arrange(sd_output)
```

## Analyzing renewable versus non-renewable energy sources

The dataset `regroup.csv` has information about which sources are considered renewable by the state of California. Use this dataset, along with your `dpylr` skills, to explore the use of renewable and non-renewable sources. Annotate what your descisions for the analysis.

Hint: Use your merge skills to merge the CA energy data with the `regroup` data. Which variable should you join by? 
